---
title: Deconvoluting causal networks
author: Gibran Hemani
date: "`r Sys.Date()`"
output: 
  pdf_document:
    keep_tex: true
---



```{r, echo=FALSE, cache=FALSE}

# http://www.mathpages.com/home/kmath198/2-1/2-1.htm
# http://videolectures.net/gpip06_mackay_gpb/

library(knitr)

read_chunk("functions.r")
source("functions.r")
opts_chunk$set(warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE)
set.seed(1234)

```

## Introduction

Growth in summary data for GWASs on phenotypes is on a steep trajectory, which is asymptoting twoards a situation where we can use two-sample MR to test 'everything against everything'. This promise, however, requires methodological advancements to address several issues including:

- Multiple testing
- Decomposing many indirect effects into a terse set of direct effects

The latter can be summarised as follows. Suppose there are five variables of interest, 1-5, and the causal relationships are

```
1 -> 2
2 -> 3
3 -> 4
4 -> 5
```

This can be depicted in graph form as in Figure 1.

```{r, fig.cap="Simulated causal relationships"}

dat1 <- init_dat(500000, 5)
dat1 <- make_edge(1, 2, -1, dat1)
dat1 <- make_edge(2, 3, -2, dat1)
dat1 <- make_edge(3, 4, 1, dat1)
dat1 <- make_edge(4, 5, 3, dat1)

plot_from_matrix(dat1$r)

```

If, however, we performed MR of `1 -> 3`, `1 -> 4`, etc, we would identify associations because they exist indirectly. Hence, after testing everything against everything our graph would look like Figure 2.

```{r, fig.cap="Empirical associations"}

res1 <- graph_mr(dat1)
plot_from_matrix(res1$b)

```

The task is to decompose the complete set of associations into the direct effects only. This has the following advantages:

- Identify direct pathways through which a particular exposure influences an outcome
- Identify instances of partial mediation, which might suggest that there are unknown variables that remain to be uncovered that mediate the path from exposure to outcome

MR for mediation has already been developed, especially for the case where there are only three phenotypic variables. Here, the direct effect of trait 1 on trait 2, $\beta_{1 \Rightarrow 2}$, is straightforward

$$
\beta_{1 \Rightarrow 2} = \beta_{1 \rightarrow 2} - \beta_{1 \rightarrow 3}\beta_{3 \rightarrow 2}
$$

With four variables it looks like:

$$
\begin{aligned}

\beta_{1 \Rightarrow 2} = \beta_{1 \rightarrow 2} & - \beta_{1 \rightarrow 3}\beta_{3 \rightarrow 4}\beta_{4 \rightarrow 2} \\
& - \beta_{1 \rightarrow 3} \beta_{3 \rightarrow 2} \\
& - \beta_{1 \rightarrow 4} \beta_{4 \rightarrow 2}

\end{aligned}
$$

```{r, eval=FALSE, fig.cap="MR for mediation"}

r <- matrix(0,6,6)
r[1,2] <- 1
r[2,3] <- 1
r[2,4] <- 1
r[4,3] <- 1
r[5,4] <- 1
r[6,2] <- 1
r[6,3] <- 1
r[6,4] <- 1

plot_from_matrix(r)

```












```{r eval=FALSE}

dat1 <- init_dat(500000, 5)
dat1 <- make_edge(1, 2, -1, dat1)
dat1 <- make_edge(1, 3, -2, dat1)
dat1 <- make_edge(2, 3, 3, dat1)
dat1 <- make_edge(3, 4, 4, dat1)
dat1 <- make_edge(3, 5, 5, dat1)
res1 <- graph_mr(dat1)

res1

plot_from_matrix(res1$b)

res1o <- get_orthogonal_graph(res1)

res1o
plot_from_matrix(round(res1o,1))
plot_from_matrix(dat1$r)
dat1
names(dat1)
dat1$r

p <- 15
dat2 <- init_dat(500000, p)
for(i in 1:(p-1))
{
	dat2 <- make_edge(i, i + 1, rnorm(1), dat2)
}
res2 <- graph_mr(dat2)
res2o <- get_orthogonal_graph(res2)

round(res2o,1)
round(dat2$r,1)
plot(res2o ~ dat2$r)



dat3 <- init_dat(500000, 4)
dat3 <- make_edge(1,2,2, dat3)
dat3 <- make_edge(2,3,3, dat3)
dat3 <- make_edge(2,4,7, dat3)
dat3 <- make_edge(1,3,4, dat3)
dat3 <- make_edge(3,4,5, dat3)
dat3 <- make_edge(1,4,6, dat3)
res3 <- graph_mr(dat3)

res3

dat3$r
round(res3$b,1)
net3 <- graph.adjacency(t(dat3$r), weighted=TRUE, mode="directed")
E(net3)$width <- abs(E(net3)$weight)
plot(net3)

dat3$r[3,1] - (res3$b[3,1] - get_prods(get_paths(1, 3, 4), round(res3$b, 1)))
dat3$r[3,2] - (res3$b[3,2] - get_prods(get_paths(2, 3, 4), round(res3$b, 1)))
dat3$r[4,2] - (res3$b[4,2] - get_prods(get_paths(2, 4, 4), round(res3$b, 1)))
dat3$r[4,3] - (res3$b[4,3] - get_prods(get_paths(3, 4, 4), round(res3$b, 1)))
dat3$r[4,1] - (res3$b[4,1] - get_prods(get_paths(1, 4, 4), round(res3$b, 1)))


# Get all the direct effect products

# Problem: within the raw matrix there are compound effects, e.g. 1-2 is direct, 1-3 is direct plus indirect, so the 1-3 element is not the orthogonal effect. To get 1-4 (via 2 and 3) we need the direct and indirect paths for all intermediate nodes. 

# Solution: Work from the network rather than the matrix. Make the network sparse, identify all paths between 1 and 4

# Network deconvolution: http://www.nature.com/nbt/journal/v31/n8/full/nbt.2635.html
# https://github.com/gidonro/Network-Deconvolution/blob/master/ND.py



dat3$r %*% solve(diag(4) - dat3$r)

dat3$r %*% (diag(4) + dat3$r + dat3$r %*% dat3$r^2 + dat3$r %*% dat3$r^2 %*% dat3$r^2)

res3$b



# Find all the paths in a sparse adjacency matrix
# Work on these instead of the full matrix
#



dat4 <- init_dat(500000, 3)
dat4 <- make_edge(1,2,2, dat4)
dat4 <- make_edge(2,3,3, dat4)
dat4 <- make_edge(1,3,4, dat4)
res4 <- graph_mr(dat4)

res4

dat4$r

get_prods(get_paths(1, 3, 3), res4$b)



round(-solve(res1$b))


r <- matrix(0,5,5)



r <- matrix(0,5,5)
diag(r) <- 1
r[2,1] <- -1
r[3,1] <- -2
r[3,2] <- 3
r[4,3] <- 4
r[5,3] <- 5

plot_from_matrix(r)





r <- matrix(0,5,5)
diag(r) <- 1
r[1,2] <- 4
r[2,3] <- -2
r[3,4] <- 8
r[4,5] <- -3

plot_from_matrix(r)


effs <- c(1:5)
n <- 500000
g <- scale(matrix(rbinom(n*5, 2, 0.5), n))
a <- g[,1] + rnorm(n)
b <- g[,2] + rnorm(n)
c <- g[,3] + rnorm(n)
d <- g[,4] + rnorm(n)
e <- g[,5] + rnorm(n)
d <- data.frame(a, b, c, d, e)

d[,2] <- d[,2] + d[,1] * 

b <- b + a * -1
c <- c + a * -2
c <- c + b * 3
d <- d + c * 4
e <- e + c * 5



round(b)
round(solve(b))
r
table(r==-round(solve(b)))



# library(glasso)
# out <- glasso(b, rho=0.5)




library(igraph)

net3 <- graph_from_incidence_matrix(abs(round(solve(b))))

dev.new()
plot(net)
dev.new()
plot(net2)
dev.new()
plot(net3)













n <- 10000
nstep <- 100
ab <- 0.1
ba <- 0.2

a <- rnorm(n)
b <- rnorm(n)

for(i in 1:nstep)
{
	a <- a + b * ba
	b <- b + a * ab
}

cor(a,b)
mean(a)
mean(b)

plot(a,b)


```

Problems:

- Cycles
- Impossible associations
- Non-gaussian effects